Relationships in the extremes and their influence on competition and coexistence

### 2022-03-07 paper/Paper.Rnw
finished readding deletions in intro

%Intro to storage effects and why they might be influenced by ATAs
It is straightforward to see, intuitively, 
why ATAs may influence one mechanism of coexistence in particular: storage effects. We review, conceptually, 
what storage effects are \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}, and we explain why ATAs may be relevant to them.
Storage effects allow for species to fluctuate over time while escaping extinction because storage effects provide a relative benefit to
whichever species is currently rare; this is the characteristic that makes storage effects a so-called \emph{stabilizing} mechanism 
of coexistence \citep{Chesson_2000}.
Storage effects can occur if three conditions are met \citep{Chesson_1994}. First, there must be positive 
covariation between the fluctuating quality of the environment (E)
for a species and the strength of competition (C) experienced by that species (called EC covariance). 
Second, growth responses to the environment must be species-specific, so that distinct species do not find the 
same environmental conditions to be optimal. Third, a buffering mechanism must exist whereby species limit the influence of competition
in times when the environment is poor. Dormant stages such as seed banks are an
example buffering mechanism \citep{Chesson_2012}. 
When present, these three factors combine so that EC covariance provides a relative benefit to whatever species are currently rare.
Importantly, storage effects can be quantified by measuring to what extent EC covariance contributes to differences between 
invader and resident growth rates \citep{Chesson_1994,Ellner_2016}.
Thus ATAs may, in principle, influence storage effects simply because storage effects hinge on EC covariation, and ATAs, when they 
occur between environment and competition variables, can be an aspect of EC covariation.

@@

%Climate change may alter ATAs of coexistence-relevant ATAs
An additional motivation for quantifying the importance of ATAs for species coexistence is that there is reason to believe climate change may
alter ATAs of environmental variables which impact coexistence, as follows.
It is well known that climate change can alter the intensity, duration, extent and frequency of extreme environmental events, including
heat waves, floods, hurricanes and fires \citep{Meehl_2004, Jentsch_2007, Ummenhofer_2017, Lyon2019, Keelings2020}.  
If extreme events are also longer lasting, then the nature of the covariation between phenologically separated environmental signals 
may be modified by changes in extreme climatic events; in particular, ATAs between such variables may be altered. 
For instance, imagine a case in which early spring precipitation influences the growth of a plant species, A,
whereas late spring precipitation influences plant species B. If these species are competitors, then EC covariation
contributing to their coexistence, or failure to coexist, can boil down to covariation 
between early and late spring precipitation. Because longer lasting weather events are more likely to contribute
to both early and late spring rainfall variables, a tendency for spring rainfall events to become both more extreme and 
longer lasting increases the degree of right-tail association between early- and late-spring rainfall. 
Likewise, a tendency for spring droughts to become both more intense and longer lasting can increase
left-tail association between these variables. As we will show in this study, changes in ATAs can then
result in different competitive outcomes between species. Although we focus here on temporal coexistence
mechanisms, ATAs between environmental variables measured in different places may also be altered 
by climate change, if more intensive climatic events are also more spatially extensive.

%make the above abstract ideas more concrete
To make the above ideas more concrete, we briefly mention examples from prior work to 
which the ideas may apply. 
\cite{Chesson_2012} provides two examples of competitive plant communities in which different species
are likely influenced differently by phenologically distict environmental variables: an Australian 
grass-forb community, and an Arizona community of desert annuals. 
The communities follow the criteria required for temporal storage effects to operate \citep{Chesson_2012}. In both 
examples, distinct species have different periods of germination, competition is influenced by resource 
consumption, and population growth is buffered by overlapping generations or via seed banks.
Temporal differences between species in germination rates suggest the possibility that 
population growth of distinct species may be driven by rainfall signals in distinct parts
of the growing season. Thus all the ingredients seem to be present by which climatic changes may
alter ATAs between environmental variables which are relevant for the coexistence
of the example species, motivating our efforts, here, to perform a detailed study of how ATAs can influence competition.

%Review of lit on changes in means and variances of climate signals and associated impacts on coexistence
Although, to our knowledge, the potential importance of climatic changes in ATAs for species coexistence has 
not been studied, the importance for coexistence of changes 
in means and variability of environmental variables has been studied in many systems. We briefly 
review some of that literature.
\cite{diPaola_2012} concluded that coexistence of drought sensitive,
flood resistant species with flood sensitive, drought resistant species in 
Mediteranean coastal swamp forests of central Italy would be jeopardized
by projected changes in mean rainfall. 
\cite{Descombes_2020} found that if warming trends promote faster expansion toward 
higher altitudes of herbivore populations than of plant populations, newly arrived
herbivores at high altitudes can inflate plant species richness by reducing dominant plant species 
in favor of subordinates. Contrastingly, \cite{Usinowicz_2021} concluded, also for
alpine plants, that upward range expansion under warming trends would be good for the growth and survival of 
all species they considered, but that changes in competition would more than offset these direct benefits. 
Numerous studies take an ``event-focused'' 
perspective \citep{Jentsch_2007}, examining how increased climatic variability, in the
form of extreme weather events, can alter competitive interactions. For instance, 
\cite{White_2001} showed that single extreme heat events can alter competition 
between C3 and C4 plants. And \cite{Parepa_2013} showed that extreme variability in soil
nutrient availability for plants, mediated by soil moisture content, can alter 
the potential for invasive species to enter an ecosystem and establish, a process that involves
the invader overcoming competition from resident species. 
Using MCT, it has been demonstrated that increased variability can promote coexistence in plant 
communities, via storage effects \citep{Adler_2006,  Angert_2009}.
Changes in variability and in extreme events are distinct from the changes in ATAs we 
consider here, because ATAs pertain to \emph{relationships} between environmental and other
variables in the extremes, which are distinct from and can be altered independently
of changes in the extremes of the univariate marginal distributions themselves \citep{Nelsen2006,Ghosh_2020_AER}.

>>>>

%Intro to storage effects and why they might be influenced by ATAs
It is straightforward to see, intuitively, 
why ATAs may influence one mechanism of coexistence in particular: storage effects. We review, conceptually, 
what storage effects are \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}, and explain why ATAs may be relevant to them.
Storage effects allow for species to fluctuate over time while escaping extinction because storage effects provide a relative benefit to
whichever species is currently rare; this is the characteristic that makes storage effects a so-called \emph{stabilizing} mechanism 
of coexistence \citep{Chesson_2000}.
Storage effects can occur if three conditions are met \citep{Chesson_1994}. First, there must be positive 
covariation between the fluctuating quality of the environment (E)
for a species and the strength of competition (C) experienced by that species (called EC covariance). 
Second, growth responses to the environment must be species-specific, so that distinct species do not find the 
same environmental conditions to be optimal. Third, a buffering mechanism must exist whereby species limit the influence of competition
in times when the environment is poor. Dormant stages such as seed banks are an
example buffering mechanism \citep{Chesson_2012}. 
When present, these three factors combine so that EC covariance provides a relative benefit to whatever species is currently rare.
Importantly, storage effects can be quantified by measuring to what extent EC covariance contributes to differences between 
invader and resident growth rates \citep{Chesson_1994,Ellner_2016}.
Thus ATAs may, in principle, influence storage effects simply because storage effects hinge on EC covariation, and ATAs, when they 
occur between environment and competition variables, can be an aspect of EC covariation.

%***DAN: Go through and make sure all citation callouts list papers in date order, oldest to newest,
%except when multiple pubs by the same first author are listed they are all together.

@@

%Climate change may alter ATAs of coexistence-relevant ATAs
An additional motivation for quantifying the importance of ATAs for species coexistence is that there is reason to believe climate change may
alter ATAs of environmental variables which impact coexistence, as follows.
It is well known that climate change can alter the intensity, duration, extent and frequency of extreme environmental events, including
heat waves, floods, hurricanes and fires \citep{Meehl_2004, Jentsch_2007, Ummenhofer_2017, Lyon2019, Keelings2020}.  
If extreme events are also longer lasting, then the nature of the covariation between phenologically separated environmental signals 
may be modified by changes in extreme climatic events; in particular, ATAs between such variables may be altered. 
For instance, imagine a case in which early spring precipitation influences the growth of a plant species, A,
whereas late spring precipitation influences plant species B. If these species are competitors, then EC covariation
contributing to their coexistence, or failure to coexist, can boil down to covariation 
between early and late spring precipitation. Because longer lasting weather events are more likely to contribute
to both early and late spring rainfall variables, a tendency for spring rainfall events to become both more extreme and 
longer lasting increases the degree of right-tail association between early- and late-spring rainfall. 
Likewise, a tendency for spring droughts to become both more intense and longer lasting can increase
left-tail association between these variables. As we will show in this study, changes in ATAs can then
result in different competitive outcomes between species. 
\cite{Chesson_2012} provides examples from an Australian grass-forb community and from an Arizona 
community of desert annuals that seem to have all the ingredients necessary to make effects of the
type we describe here a possibility.
To out knowledge, the potential importance of climatic changes in ATAs for species coexistence has 
not been studied, but the importance for coexistence of changes 
in means and variability of environmental variables has been studied in many systems
[e.g., \cite{White_2001, Adler_2006, Jentsch_2007, Angert_2009, diPaola_2012, Descombes_2020, Usinowicz_2021}].
Changes in variability and in extreme events are distinct from the changes in ATAs we 
consider here, because ATAs pertain to \emph{relationships} between environmental and other
variables in the extremes, which are distinct from and can be altered independently
of changes in the extremes of the univariate marginal distributions themselves \citep{Nelsen2006,Ghosh_2020_AER}.

%make the above abstract ideas more concrete - 
%***DAN: This para commented out for length, added a sentence about the systems to the above paragraph to take the place of this text
%To make the above ideas more concrete, we briefly mention examples from prior work to 
%which the ideas may apply. 
%\cite{Chesson_2012} provides two examples of competitive plant communities in which different species
%are likely influenced differently by phenologically distict environmental variables: an Australian 
%grass-forb community, and an Arizona community of desert annuals. 
%The communities follow the criteria required for temporal storage effects to operate \citep{Chesson_2012}. In both 
%examples, distinct species have different periods of germination, competition is influenced by resource 
%consumption, and population growth is buffered by overlapping generations or via seed banks.
%Temporal differences between species in germination rates suggest the possibility that 
%population growth of distinct species may be driven by rainfall signals in distinct parts
%of the growing season. Thus all the ingredients seem to be present by which climatic changes may
%alter ATAs between environmental variables which are relevant for the coexistence
%of the example species, motivating our efforts, here, to perform a detailed study of how ATAs can influence competition.

%Review of lit on changes in means and variances of climate signals and associated impacts on coexistence
%**DAN: this para commetned out for length, added a couple of sentences to an aboce para to substitute
%Although, to our knowledge, the potential importance of climatic changes in ATAs for species coexistence has 
%not been studied, the importance for coexistence of changes 
%in means and variability of environmental variables has been studied in many systems. We briefly 
%review some of that literature.
%\cite{diPaola_2012} concluded that coexistence of drought sensitive,
%flood resistant species with flood sensitive, drought resistant species in 
%Mediteranean coastal swamp forests of central Italy would be jeopardized
%by projected changes in mean rainfall. 
%\cite{Descombes_2020} found that if warming trends promote faster expansion toward 
%higher altitudes of herbivore populations than of plant populations, newly arrived
%herbivores at high altitudes can inflate plant species richness by reducing dominant plant species 
%in favor of subordinates. Contrastingly, \cite{Usinowicz_2021} concluded, also for
%alpine plants, that upward range expansion under warming trends would be good for the growth and survival of 
%all species they considered, but that changes in competition would more than offset these direct benefits. 
%Numerous studies take an ``event-focused'' 
%perspective \citep{Jentsch_2007}, examining how increased climatic variability, in the
%form of extreme weather events, can alter competitive interactions. For instance, 
%\cite{White_2001} showed that single extreme heat events can alter competition 
%between C3 and C4 plants. And \cite{Parepa_2013} showed that extreme variability in soil
%nutrient availability for plants, mediated by soil moisture content, can alter 
%the potential for invasive species to enter an ecosystem and establish, a process that involves
%the invader overcoming competition from resident species. 
%Using MCT, it has been demonstrated that increased variability can promote coexistence in plant 
%communities, via storage effects \citep{Adler_2006,  Angert_2009}.
